---
title: "LA_EDA"
output: html_document
date: "2024-10-26"
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
 
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

**SMART QUESTIONS**

1) Which neighborhoods in Los Angeles have experienced the most significant increases in crime rates from 2020 to the present, and are these trends consistent for the same types of crimes year over year?
2) How have theft, robbery, and homicide rates in Los Angeles changed from 2020 to the present, and which of these categories shows the highest overall density of crime during this time?
3) How does the type or frequency of crimes in Los Angeles vary by victim descent, and are there significant geographic patterns (based on latitude and longitude) associated with specific victim groups?
4) Which type of crime in Los Angeles is most frequently associated with weapon use, and what specific types of weapons are most commonly involved in those incidents over the last 2 years?
5) How have the top three most common crimes from 2020 to the present been distributed across the top five areas where they are most frequently committed in Los Angeles, and are these trends increasing or decreasing in each area from 2020 to 2024?


```{r}
crime = read.csv("Crime_Data_from_2020_to_Present.csv")
head(crime)
```

```{r}
nrow(crime)
ncol(crime)
```

The dataset contains 28 columns and 986500 rows.
1) DR_NO - Division of Records Number: Official file number.
2) Date Rptd - MM/DD/YYYY
3) DATE OCC - MM/DD/YYYY
4) TIME OCC - In 24 hour military time
5) AREA - The LAPD has 21 Community Police Stations referred to as Geographic Areas within the department. These Geographic Areas are sequentially numbered from 1-21.
6) AREA NAME - The 21 Geographic Areas or Patrol Divisions are also given a name designation that references a landmark or the surrounding community that it is responsible for.
7) Crm Cd - Indicates the crime committed. Crime Code 1 is the primary and most serious one. Crime Code 2, 3, and 4 are respectively less serious offenses. Lower crime class numbers are more serious.
8) Crm Desc - Indicates the crime description
9) Vict Age - Age of victim
10) Vict Sex - 	F : Female M : Male X : Unknown
11) Vict Descent - 	Descent Code: A - Other Asian B - Black C - Chinese D - Cambodian F - Filipino G - Guamanian H - Hispanic/Latin/Mexican I - American Indian/Alaskan Native J - Japanese K - Korean L - Laotian O - Other P - Pacific Islander S - Samoan U - Hawaiian V - Vietnamese W - White X - Unknown Z - Asian Indian
12) Weapon Desc - Defines the Weapon Used Code provided.
13) Location - Street address of crime incident rounded to the nearest hundred block to maintain anonymity.
14) LAT - Latitude
15) LON - Longtitude


```{r}
na_count <- colSums(is.na(crime))
print("\nNA Count per Column:")
print(na_count)


```

```{r}
new_crime <- crime[, colSums(is.na(crime)) == 0]
head(new_crime)

nrow(new_crime)
ncol(new_crime)
```

```{r}
# Feature Selection based on Smart Questions

cols_to_remove <- c(
   "Mocodes", "Rpt.Dist.No", "Part.1.2", 
  "Premis_cd","Premis.Desc", "Status", "Status.Desc","Cross.Street"
)

# Drop the specified columns
crime_data <- new_crime[, !(names(new_crime) %in% cols_to_remove)]


print("Data after removing unnecessary columns:")
print(names(crime_data))
```

```{r}

library(dplyr)
library(ggplot2)
library(lubridate)

# Ensure date is in correct format
crime_data$DATE.OCC <- as.Date(crime_data$DATE.OCC, format = "%m/%d/%Y")


# Identify top 3 crimes overall from 2020 to present
top_crimes <- crime_data %>%
  filter(year(DATE.OCC) >= 2020) %>%
  group_by(Crm.Cd, Crm.Cd.Desc) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  head(3)
crime_summary <- crime_data %>%
  filter(year(DATE.OCC) >= 2020) %>%
  group_by(AREA.NAME) %>%
  summarise(
    total_crime = n(),
    top3_crime = sum(Crm.Cd %in% top_crimes$Crm.Cd)
  ) %>%
  mutate(top3_pct = top3_crime / total_crime * 100)
top_areas <- crime_summary %>%
  arrange(desc(top3_crime)) %>%
  head(5)
# Identify top 5 areas with highest counts of top 3 crimes


# Filter data for top 3 crimes from 2020 onwards
crime_filtered <- crime_data %>%
  filter(year(DATE.OCC) >= 2020, Crm.Cd %in% top_crimes$Crm.Cd)
```



Question 5:
How have the top three most common crimes from 2020 to the present been distributed across the top five areas where they are most frequently committed in Los Angeles, and are these trends increasing or decreasing in each area from 2020 to 2024?

```{r}

library(dplyr)
library(ggplot2)
library(lubridate)

# Ensure date is in correct format
crime_data$DATE.OCC <- as.Date(crime_data$DATE.OCC, format = "%m/%d/%Y")

# Calculate crime counts by year for top 3 crimes in top 5 areas
crime_counts <- crime_data %>%
  filter(year(DATE.OCC) >= 2020, 
         AREA.NAME %in% top_areas$AREA.NAME, 
         Crm.Cd %in% top_crimes$Crm.Cd) %>%
  group_by(AREA.NAME, Year = year(DATE.OCC), Crm.Cd.Desc) %>%
  summarise(crime_count = n()) %>%
  ungroup()

# Calculate percent change from 2020 benchmark
crime_change <- crime_counts %>%
  group_by(AREA.NAME, Crm.Cd.Desc) %>%
  mutate(
    benchmark_2020 = crime_count[Year == 2020],
    percent_change = (crime_count - benchmark_2020) / benchmark_2020 * 100
  ) %>%
  ungroup()

# Create a line graph showing percent change from 2020 benchmark for each area
p <- ggplot(crime_change, aes(x = Year, y = percent_change, color = Crm.Cd.Desc, group = Crm.Cd.Desc)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  facet_wrap(~ AREA.NAME, ncol = 2) +
  theme_minimal() +
  labs(title = "Percent Change in Top 3 Crimes from 2020 Benchmark - Top 5 Areas",
       x = "Year",
       y = "Percent Change from 2020",
       color = "Crime Type") +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, size = 10)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

print(p)

# Print crime counts and percent changes for each area
print(crime_change)
```



Question 1

Which neighborhoods in Los Angeles have experienced the most significant increases in crime rates from 2020 to 2023 with particular attention to demographic factors (race and sex of victims), and what insights can be drawn from Area3's significant increase in crimes?

```{r}

library(dplyr)
library(tidyr)
library(knitr)
library(lubridate)

# Assuming crime_data is your original dataset and DATE.OCC is a date column
crime_data_filtered <- crime_data %>%
  filter(year(DATE.OCC) < 2024)

# Create a summary table with year-over-year percentage changes
crime_summary <- crime_data_filtered %>%
  mutate(Year = year(DATE.OCC)) %>%
  filter(Year >= 2020) %>%
  group_by(AREA.NAME, Year) %>%
  summarise(crime_count = n(), .groups = "drop") %>%
  group_by(AREA.NAME) %>%
  mutate(
    previous_year_count = lag(crime_count),
    yoy_percent_change = (crime_count - previous_year_count) / previous_year_count * 100
  ) %>%
  ungroup()

# Create a wide format table
crime_table <- crime_summary %>%
  select(AREA.NAME, Year, yoy_percent_change) %>%
  pivot_wider(
    names_from = Year,
    values_from = yoy_percent_change,
    names_prefix = "percent_change_"
  )

# Print the table
kable(crime_table, format = "markdown", digits = 2)

```

```{r}

library(dplyr)
library(lubridate)

# Assuming crime_data is your original dataset and DATE.OCC is a date column
crime_summary <- crime_data %>%
  mutate(Year = year(DATE.OCC)) %>%
  filter(Year %in% c(2020, 2023)) %>%
  group_by(AREA.NAME, Year) %>%
  summarise(crime_count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Year, values_from = crime_count) %>%
  mutate(percent_change = (`2023` - `2020`) / `2020` * 100) %>%
  arrange(desc(percent_change))

# Print the summary
print(crime_summary)



```


```{r}
library(dplyr)
library(lubridate)

# Calculate crime counts and percent change by area and crime type
crime_analysis <- crime_data %>%
  filter(year(DATE.OCC) %in% c(2020, 2023), 
         AREA.NAME %in% top_areas$AREA.NAME) %>%
  group_by(AREA.NAME, Crm.Cd.Desc, Year = year(DATE.OCC)) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Year, 
              values_from = count, 
              names_prefix = "count_") %>%
  mutate(
    percent_change = ((count_2023 - count_2020) / count_2020) * 100,
    absolute_change = count_2023 - count_2020
  ) %>%
  filter(!is.na(percent_change)) %>%
  arrange(AREA.NAME, desc(percent_change))

# Get top 3 most increased crimes for each area
top_crimes_by_area <- crime_analysis %>%
  group_by(AREA.NAME) %>%
  filter(percent_change > 0) %>%  # Only include increases
  top_n(3, percent_change) %>%
  arrange(AREA.NAME, desc(percent_change))

# Print results
print(top_crimes_by_area)

```


```{r}
library(dplyr)
library(ggplot2)
library(lubridate)

# Calculate crime changes by Area, Crime Type, Sex, and Race
crime_analysis <- crime_data %>%
  filter(year(DATE.OCC) %in% c(2020, 2023)) %>%
  group_by(AREA.NAME, Crm.Cd.Desc, Vict.Sex, Vict.Descent, Year = year(DATE.OCC)) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Year, 
              values_from = count, 
              names_prefix = "count_") %>%
  mutate(
    percent_change = ((count_2023 - count_2020) / count_2020) * 100,
    absolute_change = count_2023 - count_2020
  ) %>%
  filter(!is.na(percent_change))

# Create multiple visualizations

# 1. Crime Change by Area and Sex
p1 <- ggplot(crime_analysis, 
       aes(x = AREA.NAME, y = percent_change, fill = Vict.Sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Crime Change by Area and Victim Sex (2020-2023)",
       x = "Area",
       y = "Percent Change",
       fill = "Victim Sex")

# 2. Crime Change by Area and Race/Descent
p2 <- ggplot(crime_analysis, 
       aes(x = AREA.NAME, y = percent_change, fill = Vict.Descent)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Crime Change by Area and Victim Descent (2020-2023)",
       x = "Area",
       y = "Percent Change",
       fill = "Victim Descent")




# Print all plots
print(p1)
print(p2)

# Create summary table
summary_table <- crime_analysis %>%
  group_by(AREA.NAME, Vict.Sex, Vict.Descent) %>%
  summarise(
    avg_percent_change = mean(percent_change, na.rm = TRUE),
    total_crimes_2020 = sum(count_2020, na.rm = TRUE),
    total_crimes_2023 = sum(count_2023, na.rm = TRUE)
  ) %>%
  arrange(desc(avg_percent_change))

print(summary_table)


```
```{r}
head(crime$Date_Reported)
```

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)

# Display the structure to understand the format of the Date column
str(crime)

# Print the first few entries of the Date_Reported column to check its format
print(head(crime$Date_Reported))

# Assuming the date is in a column named 'Date_Reported' and format is 'mm/dd/yyyy'
# You might need to adjust this if the format does not match
crime$Date.Rptd <- as.Date(crime$Date.Rptd, format="%m/%d/%Y")

# Check if there were any conversion errors
sum(is.na(crime$Date.Rptd))

# Extract Year, Month, and Day from the Date_Reported
crime$Year <- year(crime$Date.Rptd)
crime$Month <- month(crime$Date.Rptd)
crime$Day <- day(crime$Date.Rptd)

# Summarize total crimes by month and year
monthly_crime_summary <- crime %>%
  group_by(Year, Month) %>%
  summarise(Total_Crimes = n(), .groups = 'drop')

# Plot total crimes by month and year
mcs <- ggplot(monthly_crime_summary, aes(x = Month, y = Total_Crimes, group = Year, color = as.factor(Year))) +
  geom_line() +
  labs(title = "Monthly Crime Trends by Year", x = "Month", y = "Total Crimes") +
  scale_x_continuous(breaks = 1:12, labels = month.abb)

ggplotly(mcs)

# Analyzing crime types during November and December across all years
holiday_crimes <- crime %>%
  filter(Month %in% c(11, 12)) %>%
  group_by(Year, Crm.Cd.Desc) %>%
  summarise(Total_Crimes = n(), .groups = 'drop') %>%
  arrange(Year, desc(Total_Crimes))

# Print the top crime types during holidays
print(holiday_crimes)

# Plotting top crime types during holiday months
# Plot with improved label visibility
top_crime_types <- holiday_crimes %>%
  group_by(Crm.Cd.Desc) %>%
  summarise(Total_Crimes = sum(Total_Crimes), .groups = 'drop') %>%
  top_n(20, Total_Crimes)

# Filter the main data to include only these top crime types
filtered_crimes <- holiday_crimes %>%
  filter(Crm.Cd.Desc %in% top_crime_types$Crm.Cd.Desc)

# Plot with top crime types
fc <- ggplot(filtered_crimes, aes(x = reorder(Crm.Cd.Desc, -Total_Crimes), y = Total_Crimes, fill = as.factor(Year))) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Top 20 Crime Types During Holidays", x = "Crime Type", y = "Total Crimes") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip()  # This flips the axes to make the plot horizontal

ggplotly(fc)
```

Seasonal Trends: The first plot, displaying monthly crime trends by year, shows distinct seasonal fluctuations. Typically, crime rates appear to increase during certain months, which might correlate with seasonal activities or societal patterns. Identifying these peaks can help in planning better law enforcement deployment during high-risk times.

Top Crimes During Holidays: The list and plot of top crime types during November and December for each year reveal that certain types of crimes, such as vehicle theft, burglary, and assault, consistently rank high during the holiday seasons. This pattern suggests a need for targeted preventive measures during these periods when specific crimes spike, possibly due to the increased opportunity (with many homes potentially left empty and more retail activity).

```{r}
# Annual trend of total crimes during November and December
annual_holiday_crime_trends <- holiday_crimes %>%
  group_by(Year) %>%
  summarise(Total_Crimes = sum(Total_Crimes), .groups = 'drop')

ahc <- ggplot(annual_holiday_crime_trends, aes(x = Year, y = Total_Crimes)) +
  geom_line(group=1, color="blue") +
  geom_point(color="red") +
  labs(title = "Annual Holiday Crime Trends", x = "Year", y = "Total Crimes")

ggplotly(ahc)

# Add a day of the week column
crime$Day_of_Week <- wday(crime$Date.Rptd, label = TRUE, abbr = FALSE)

# Summarize crimes by day of the week during the holiday months
weekday_holiday_crime_summary <- crime %>%
  filter(Month %in% c(11, 12)) %>%
  group_by(Day_of_Week) %>%
  summarise(Total_Crimes = n(), .groups = 'drop') %>%
  arrange(Day_of_Week)

dow <- ggplot(weekday_holiday_crime_summary, aes(x = Day_of_Week, y = Total_Crimes, fill = Day_of_Week)) +
  geom_bar(stat = "identity") +
  labs(title = "Crime Distribution by Day of the Week During Holidays", x = "Day of the Week", y = "Total Crimes")

ggplotly(dow)
```

Visual Accessibility: The bar chart displaying top 20 crime types during the holidays uses a horizontal layout, making it easier to read and compare the frequency of different crimes. This visualization helps stakeholders quickly identify the most prevalent crimes during the holiday seasons, which could inform public awareness campaigns and policing strategies.

Detailed Examination: The detailed breakdown by crime type, year, and month provides a granular view of crime trends, offering insights not only into the ‘what’ and ‘when’ but potentially the ‘why’ of crime patterns. For instance, understanding that thefts increase during November and December could be linked to the higher volume of shopping and unattended properties during the holiday season.


