---
title: "L.A. Crime Scene Investigation: Data-Driven Detectives"
author: "
          Abhinaysai Kamineni
          Lasya Raghavendra
          Neeraj Magadum
          Aakash Hariharan 
          Amogh Ramagiri
"
date: "2024-12-02"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
 
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 

```

# SMART QUESTIONS

1. How effectively can crime prediction models utilize historical data on the top three most common crimes in Los Angeles to forecast their distribution across high-frequency areas?

2. How can we analyze the increase in crime rates across Los Angeles neighborhoods from 2020 to 2023, using demographic factors such as the race and sex of victims to identify key patterns and potential underlying causes contributing to the rise in criminal activity?

3. How effectively can crime prediction models utilize victim descent, spatial data, and temporal patterns to forecast the likelihood and type of criminal incidents in different areas of Los Angeles?

4. How well can we use past data on holiday-season crimes in Los Angeles to predict where and how often the top three common crimes might happen in busy areas during November and December?

5. How can we predict the likelihood of different types of weapons being used in crimes (such as  theft, robbery, and homicide) in Los Angeles over the next 24 months, using crime data from the last 4 years?


```{r 1, echo=T, results='markup'}
crime = read.csv("Crime_Data.csv")
head(crime)
```
## Date Analysis and Cleaning 0.1
```{r 2, echo=T, results='markup'}
nrow(crime)
ncol(crime)
```

The dataset contains 28 columns and 986500 rows.

1) DR_NO - Division of Records Number: Official file number.
2) Date Rptd - MM/DD/YYYY
3) DATE OCC - MM/DD/YYYY
4) TIME OCC - In 24 hour military time
5) AREA - The LAPD has 21 Community Police Stations referred to as Geographic Areas within the department. These Geographic Areas are sequentially numbered from 1-21.
6) AREA NAME - The 21 Geographic Areas or Patrol Divisions are also given a name designation that references a landmark or the surrounding community that it is responsible for.
7) Crm Cd - Indicates the crime committed. Crime Code 1 is the primary and most serious one. Crime Code 2, 3, and 4 are respectively less serious offenses. Lower crime class numbers are more serious.
8) Crm Desc - Indicates the crime description
9) Vict Age - Age of victim
10) Vict Sex - 	F : Female M : Male X : Unknown
11) Vict Descent - 	Descent Code: A - Other Asian B - Black C - Chinese D - Cambodian F - Filipino G - Guamanian H - Hispanic/Latin/Mexican I - American Indian/Alaskan Native J - Japanese K - Korean L - Laotian O - Other P - Pacific Islander S - Samoan U - Hawaiian V - Vietnamese W - White X - Unknown Z - Asian Indian
12) Weapon Desc - Defines the Weapon Used Code provided.
13) Location - Street address of crime incident rounded to the nearest hundred block to maintain anonymity.
14) LAT - Latitude
15) LON - Longtitude


```{r 3, echo=T, results='markup'}
na_count <- colSums(is.na(crime))
print("\nNA Count per Column:")
print(na_count)
```

```{r 4, echo=T, results='markup'}
new_crime <- crime[, colSums(is.na(crime)) == 0]
head(new_crime)

nrow(new_crime)
ncol(new_crime)
```
## Removing Unwanted Columns 0.2
```{r 5, echo=T, results='markup'}
# Feature Selection based on Smart Questions

cols_to_remove <- c(
   "Mocodes", "Rpt.Dist.No", "Part.1.2", 
  "Premis_cd","Premis.Desc", "Status", "Status.Desc","Cross.Street"
)

# Drop the specified columns
crime_data <- new_crime[, !(names(new_crime) %in% cols_to_remove)]


print("Data after removing unnecessary columns:")
print(names(crime_data))
```

```{r 6, echo=T, results='markup'}
# Date/Time Processing.
crime_data$Date_Rptd <- as.POSIXct(crime_data$Date.Rptd, format="%m/%d/%Y")
crime_data$Date_Occ <- as.POSIXct(crime_data$DATE.OCC, format="%m/%d/%Y")

crime_data$day_of_week <- weekdays(crime_data$Date_Occ)
crime_data$hour <- floor(crime_data$TIME.OCC/100)
crime_data$time_to_report <- as.numeric(difftime(crime_data$Date_Rptd, crime_data$Date_Occ, units="days"))
```

```{r 7, echo=T, results='markup'}
# Feature Level Engineering of Preprocessing.
crime_data$has_weapon <- ifelse(crime_data$Weapon.Desc == "STRONG-ARM (HANDS, FIST, FEET OR BODILY FORCE)", 0, 1)

crime_data$age_group <- cut(crime_data$Vict.Age, 
                           breaks=c(0, 18, 30, 50, 70, 100),
                           labels=c("Minor", "Young Adult", "Adult", "Senior", "Elderly"))

crime_data$time_period <- cut(crime_data$hour, 
                             breaks=c(0, 6, 12, 18, 24),
                             labels=c("Night", "Morning", "Afternoon", "Evening"))
```

```{r 8, echo=T, results='markup'}
# Categorical Encoding
#install.packages("caret")
library(caret)
dummy_vars <- dummyVars(" ~ AREA.NAME + Vict.Sex + Crm.Cd.Desc", data = crime_data)
categorical_encoded <- predict(dummy_vars, newdata = crime_data)
crime_data_encoded <- cbind(crime_data, categorical_encoded)
```


```{r 9, echo=T, results='markup'}
# Outlier Detection
q1_age <- quantile(crime_data$Vict.Age, 0.25)
q3_age <- quantile(crime_data$Vict.Age, 0.75)
iqr_age <- q3_age - q1_age
crime_data$age_outlier <- crime_data$Vict.Age < (q1_age - 1.5 * iqr_age) | 
                         crime_data$Vict.Age > (q3_age + 1.5 * iqr_age)

crime_data$coord_outlier <- crime_data$LAT < 33.7037 | crime_data$LAT > 34.3373 |
                           crime_data$LON < -118.6682 | crime_data$LON > -118.1553

crime_data$reporting_outlier <- crime_data$time_to_report > 365

crime_data_clean <- crime_data[!crime_data$age_outlier & 
                              !crime_data$coord_outlier & 
                              !crime_data$reporting_outlier, ]
```

## Data Cleaning 0.2
```{r 10, echo=T, results='markup'}
la_crime <- crime %>%
  filter(!is.na(Weapon.Used.Cd))

cols_to_remove <- c(
   "Mocodes", "Rpt.Dist.No", "Part.1.2", "Crm.Cd.2","Crm.Cd.3","Crm.Cd.4",
  "Premis_cd","Premis.Desc", "Status", "Status.Desc","Cross.Street"
)

# Drop the specified columns
la_crime<- la_crime[, !(names(la_crime) %in% cols_to_remove)]

la_crime <- la_crime %>% rename(
  Division_NO = 'DR_NO',
  Date_Reported = 'Date.Rptd',
  Date_Occurred = 'DATE.OCC',
  Time_Occurred = 'TIME.OCC',
  Area_Code = 'AREA',
  Area_Name = 'AREA.NAME',
  Crime_Code = 'Crm.Cd',
  Crime_Description = 'Crm.Cd.Desc',
  Weapons_Used='Weapon.Used.Cd',    
  Weapons_Description='Weapon.Desc',
  Victim_Age = 'Vict.Age',
  Victim_Sex = 'Vict.Sex',
  Victim_Descent = 'Vict.Descent',
  Crime_Code_1 = 'Crm.Cd',
  Location = 'LOCATION',
  Latitude = 'LAT',
  Longitude = 'LON'
)

la_crime <- la_crime %>%
  filter(Victim_Sex != "H")

```

